\textbf{ Spectrophotometric parallaxes with \project{The Cygnet}:\\ Accurate distances for luminous red-giant stars }

### 2018-08-02 paper/sp.tex
notes to self
citations and acks

% To-do items
% -----------
% - draft introduction
% - draft discussion
% - make sure we address the sensibleness and generality of the linear model for the photometric space
% - address the non-convexity of the model (see twitter convo with @andy_l_jones).
% - Coryn's point that we should carefully explain how our model is geometric.
% - related: HWR's point that this is standardizeable candles, like the Ia SNe results.
% - related: Put in an explanation of WHY this works so well.
% - related: Put in an explanation of how it is that we aren't really modeling luminosity or distance but really just generating parallaxes.
% - does the model need a cute name, like The Swan?
% - we need a Facilities list / tags
% - we need a Software list / tags
% - check with WISE experts that we have the correct WISE photometry

@@

% text macros
\newcommand{\code}[1]{\texttt{\detokenize{#1}}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\wise}{\project{\acronym{WISE}}}
\newcommand{\zmass}{\project{\acronym{2MASS}}}
\newcommand{\sdssiv}{\project{\acronym{SDSS-IV}}}
\newcommand{\sdssv}{\project{\acronym{SDSS-V}}}

@@

What follows is a regression:
We use \gaia\ measurements of parallax to learn relationships between 
the spectroscopic and photometric properties of stars and their parallaxes.
We then use that learned model to estimate improved parallaxes for stars with
poor (or no) \gaia\ measurements.
This regression is very sensitive to certain aspects of the data or experimental
design..
One is that we will do far better with the regression if we have an accurate
model for noise process in the \gaia\ data.
For this reason, we make use of a justifiable likelihood function for the
\gaia\ astrometry (CITE HOGG).
Another is that regressions can be strongly biased if we have a bad selection function
or data censoring.
This means that we cannot apply cuts to the \gaia\ astrometry or signal-to-noise
unless those cuts are correctly accounted for in our generative model for the
data sample.
Below, we will make no such cuts. That means that our training step has the odd
property that much of the training data is low in signal-to-noise and many of the
training parallax measurements are even negative!
The fact that we are using a justified likelihood function ensures that these
data will only affect the model in sensible and justifiable ways.
The alternative---cutting to high signal-to-noise data---is the more traditional
approach, but it leads to biased inferences (unless the cuts are included correctly
in the likelihood function).
The method in which we make no cuts and keep an umodified, justified likelihood function
is unbiased, and simple.

@@

Our position is that a methodological technique is correct inasmuch as
it delivers correct or best results \emph{under its particular assumptions}.
In that spirit, we present here the assumptions of the method
explicitly.
We will return to these assumptions in the Discussion section below,
to address the costs and benefits of each in more depth.
\begin{description}
\item[features] Perhaps our most fundamental assumption is that the parallax
of a star can be predicted from the features we provide, which are
the full set of (pseudo-continuum-normalized) pixels from the \apogee\ spectrum,
plus the $G$ and $({BP}-{RP})$ photometry from \gaia,
plus the $J$, $H$, and $K_s$ photometry from \zmass,
plus the $W_1$ and $W_2$ photometry from \wise.
That is, we assume that these spectrophotometric \emph{features} are sufficient
to predict the parallax in the face of variations in stellar
age, evolutionary phase, composition, and other parameters, and also interstellar
extinction.

@@

In addition, we assume that many entries in the $D$-vector will be zero
(the ``sparsity'' assumption).
In order to encourage this, we optimize not $\chi^2$ but a regularized objective function
\begin{eqnarray}
\hat{\theta} &\leftarrow& \argmin_{\theta}\left[\frac{1}{2}\,\chi^2(\theta) + ||\lambda\ast\theta||_1^1\right]
\quad ,
\end{eqnarray}
where
$\lambda$ is a $D$-vector of regularization parameters (such that we can regularize
different parts of the $x_n$ vectors differently),
the ``$a\ast b$'' operator indicates element-wise multiplication of vectors $a$ and $b$,
and $||y||_1^1$ is the L1-norm or sum of absolute values of the components of $y$.
This kind of regularization adds a convex term to the optimization and leads to
sparse optima.
We will set the value $\lambda$ of the $D$-vector of regularization strengths by a
very coarse cross-validation.

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\quad ,
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according to (HOGG WHAT EQUATION),
and
$x_m$ is the feature $D$-vector for star $m$.
Thus the model can be trained on a training set of stars and applied to make
predictions for a validation set or a test set.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

@@

Because the \apogee\ target selection is based on \zmass, every \apogee\ star
also comes with \zmass\ (HOGG CITE) photometry.
That is, for each star $n$,
we have \zmass\ near-infrared photometry $J_n$, $H_n$, and $K_n$.

@@

\wise\ Acknowledgement?

\project{\acronym{SDSS-IV}} ack.

\begin{thebibliography}{}
\bibitem[Astropy Collaboration et al.(2013)]{2013A&A...558A..33A} Astropy Collaboration, Robitaille, T.~P., Tollerud, E.~J., et al.\ 2013, \aap, 558, A33 
\bibitem[Bertin \& Arnouts(1996)]{1996A&AS..117..393B} Bertin, E., \& Arnouts, S.\ 1996, \aaps, 117, 393 
\bibitem[Corrales(2015)]{2015ApJ...805...23C} Corrales, L.\ 2015, \apj, 805, 23
\bibitem[Ferland et al.(2013)]{2013RMxAA..49..137F} Ferland, G.~J., Porter, R.~L., van Hoof, P.~A.~M., et al.\ 2013, \rmxaa, 49, 137
\bibitem[Hanisch \& Biemesderfer(1989)]{1989BAAS...21..780H} Hanisch, R.~J., \& Biemesderfer, C.~D.\ 1989, \baas, 21, 780 
\bibitem[Lamport(1994)]{lamport94} Lamport, L. 1994, LaTeX: A Document Preparation System, 2nd Edition (Boston, Addison-Wesley Professional)
\bibitem[Schwarz et al.(2011)]{2011ApJS..197...31S} Schwarz, G.~J., Ness, J.-U., Osborne, J.~P., et al.\ 2011, \apjs, 197, 31  
\bibitem[Vogt et al.(2014)]{2014ApJ...793..127V} Vogt, F.~P.~A., Dopita, M.~A., Kewley, L.~J., et al.\ 2014, \apj, 793, 127  
\end{thebibliography}

%% This command is needed to show the entire author+affilation list when
%% the collaboration and author truncation commands are used.  It has to
%% go at the end of the manuscript.
%\allauthors

%% Include this line if you are using the \added, \replaced, \deleted
%% commands to see a summary list of all changes at the end of the article.
%\listofchanges

>>>>

% To-do items
% -----------
% - draft discussion
% - make sure we address the sensibleness and generality of the linear model for the photometric space
% - address the non-convexity of the model (see twitter convo with @andy_l_jones). Plus optimization issues.
% - Coryn's point that we should carefully explain how our model is geometric.
% - related: HWR's point that this is standardizeable candles, like the Ia SNe results.
% - related: Put in an explanation of WHY this works so well.
% - related: Put in an explanation of how it is that we aren't really modeling luminosity or distance but really just generating parallaxes.
% - does the model need a cute name, like The Swan?
% - we need a Facilities list / tags
% - we need a Software list / tags
% - check with WISE experts that we have the correct WISE photometry

@@

% text macros
\newcommand{\equationname}{equation}
\newcommand{\code}[1]{\texttt{\detokenize{#1}}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\etal}{\foreign{et al.}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\wise}{\project{\acronym{WISE}}}
\newcommand{\zmass}{\project{\acronym{2MASS}}}
\newcommand{\sdssiv}{\project{\acronym{SDSS-IV}}}
\newcommand{\sdssv}{\project{\acronym{SDSS-V}}}

@@

What follows is a regression:
We use \gaia\ measurements of parallax to learn relationships between 
the spectroscopic and photometric properties of stars and their parallaxes.
We then use that learned model to estimate improved parallaxes for stars with
poor (or no) \gaia\ measurements.
This regression is very sensitive to certain aspects of the data or experimental
design..
One is that we will do far better with the regression if we have an accurate
model for noise process in the \gaia\ data.
For this reason, we make use of a justifiable likelihood function for the
\gaia\ astrometry (\citealt{gaialf}).
Another is that regressions can be strongly biased if we have a bad selection function
or data censoring.
This means that we cannot apply cuts to the \gaia\ astrometry or signal-to-noise
unless those cuts are correctly accounted for in our generative model for the
data sample.
Below, we will make no such cuts. That means that our training step has the odd
property that much of the training data is low in signal-to-noise and many of the
training parallax measurements are even negative!
The fact that we are using a justified likelihood function ensures that these
data will only affect the model in sensible and justifiable ways.
The alternative---cutting to high signal-to-noise data---is the more traditional
approach, but it leads to biased inferences (unless the cuts are included correctly
in the likelihood function).
The method in which we make no cuts and keep an umodified, justified likelihood function
is unbiased, and simple.

@@

Our position is that a methodological technique is correct inasmuch as
it delivers correct or best results \emph{under its particular assumptions}.
In that spirit, we present here the assumptions of the method
explicitly.
We will return to these assumptions in the Discussion section below,
to address the costs and benefits of each in more depth.
\begin{description}
\item[features] Perhaps our most fundamental assumption is that the parallax
of a star can be predicted from the features we provide, which are
the full set of (pseudo-continuum-normalized) pixels from the \apogee\ spectrum,
plus the $G$ and $({BP}-{RP})$ photometry from \gaia,
plus the $J$, $H$, and $K_s$ photometry from \zmass (\citealt{zmass}),
plus the $W_1$ and $W_2$ photometry from \wise.
That is, we assume that these spectrophotometric \emph{features} are sufficient
to predict the parallax in the face of variations in stellar
age, evolutionary phase, composition, and other parameters, and also interstellar
extinction.

@@

In addition, we assume that many entries in the $D$-vector will be zero
(the ``sparsity'' assumption).
In order to encourage this, we optimize not $\chi^2$ but a regularized objective function
\begin{eqnarray}
\hat{\theta} &\leftarrow& \argmin_{\theta}\left[\frac{1}{2}\,\chi^2(\theta) + ||\lambda\ast\theta||_1^1\right]
\label{eq:objective}\quad ,
\end{eqnarray}
where
$\lambda$ is a $D$-vector of regularization parameters (such that we can regularize
different parts of the $x_n$ vectors differently),
the ``$a\ast b$'' operator indicates element-wise multiplication of vectors $a$ and $b$,
and $||y||_1^1$ is the L1-norm or sum of absolute values of the components of $y$.
This kind of regularization adds a convex term to the optimization and leads to
sparse optima.
We will set the value $\lambda$ of the $D$-vector of regularization strengths by a
very coarse cross-validation.

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\quad ,
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according
to \equationname~(\ref{eq:objective}),
and
$x_m$ is the feature $D$-vector for star $m$.
Thus the model can be trained on a training set of stars and applied to make
predictions for a validation set or a test set.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

@@

Because the \apogee\ target selection is based on \zmass, every \apogee\ star
also comes with \zmass\ (\citealt{zmass}) photometry.
That is, for each star $n$,
we have \zmass\ near-infrared photometry $J_n$, $H_n$, and $K_n$.

@@

This publication makes use of data products from the Two Micron All Sky Survey, which is a joint project of the University of Massachusetts and the Infrared Processing and Analysis Center/California Institute of Technology, funded by the National Aeronautics and Space Administration and the National Science Foundation.

This publication makes use of data products from the Wide-field Infrared Survey Explorer, which is a joint project of the University of California, Los Angeles, and the Jet Propulsion Laboratory/California Institute of Technology, funded by the National Aeronautics and Space Administration.

Funding for the \project{Sloan Digital Sky Survey IV} has been provided by the Alfred P. Sloan Foundation, the U.S Department of Energy Office of Science, and the Participating Institutions. \sdssiv\ acknowledges
support and resources from the Center for High-Performance Computing at
the University of Utah. The \project{\acronym{SDSS}} web site is www.sdss.org.

\sdssiv\ is managed by the Astrophysical Research Consortium for the 
Participating Institutions of the \project{\acronym{SDSS}} Collaboration including the 
Brazilian Participation Group, the Carnegie Institution for Science, 
Carnegie Mellon University, the Chilean Participation Group, the French Participation Group, Harvard-Smithsonian Center for Astrophysics, 
Instituto de Astrof\'isica de Canarias, The Johns Hopkins University, 
Kavli Institute for the Physics and Mathematics of the Universe / 
University of Tokyo, the Korean Participation Group, Lawrence Berkeley National Laboratory, 
Leibniz Institut f\"ur Astrophysik Potsdam,  
Max-Planck-Institut f\"ur Astronomie (Heidelberg), 
Max-Planck-Institut f\"ur Astrophysik (Garching), 
Max-Planck-Institut f\"ur Extraterrestrische Physik, 
National Astronomical Observatories of China, New Mexico State University, 
New York University, University of Notre Dame, 
Observat\'ario Nacional~/~\acronym{MCTI}, The Ohio State University, 
Pennsylvania State University, Shanghai Astronomical Observatory, 
United Kingdom Participation Group,
Universidad Nacional Aut\'onoma de M\'exico, University of Arizona, 
University of Colorado Boulder, University of Oxford, University of Portsmouth, 
University of Utah, University of Virginia, University of Washington, University of Wisconsin, 
Vanderbilt University, and Yale University.

\begin{thebibliography}{}
\bibitem[Hogg(2018)]{gaialf} Hogg, D.~W 2018, arXiv:1804.07766 
\bibitem[Skrutskie \etal(2006)]{zmass} Skrutskie, M.~F., Cutri, R.~M., Stiening, R., et al.\ 2006, \aj, 131, 1163 
\end{thebibliography}
